\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{tikz-cd}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}

\DeclareUnicodeCharacter{03C6}{\ensuremath{\varphi}}
\DeclareUnicodeCharacter{03A6}{\ensuremath{\Phi}}
\DeclareUnicodeCharacter{03B3}{\ensuremath{\gamma}}
\DeclareUnicodeCharacter{03BC}{\ensuremath{\mu}}
\DeclareUnicodeCharacter{03B7}{\ensuremath{\eta}}
\DeclareUnicodeCharacter{03C4}{\ensuremath{\tau}}
\DeclareUnicodeCharacter{2243}{\ensuremath{\simeq}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\lstdefinelanguage{lean}{
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{gray},
    morekeywords={import, namespace, open, universe, abbrev, class, extends, where, instance, section, variable, theorem, def, by, intros, rfl, ext, dsimp, erw, rw, change, congr, exact, have, fun, let},
    morecomment=[l]{--},
    morecomment=[s]{/-}{-/},
    literate=
        {⥤}{{$\rightsquigarrow$}}1
        {⋙}{{$\ggg$}}1
        {⟶}{{$\longrightarrow$}}1
        {𝟭}{{$\mathbf{1}$}}1
        {𝟙}{{$\text{id}$}}1
        {γμ}{{$\gamma\mu$}}2
        {γη}{{$\gamma\eta$}}2
        {γ}{{$\gamma$}}1
        {μ}{{$\mu$}}1
        {η}{{$\eta$}}1
        {τ}{{$\tau$}}1
        {φ}{{$\varphi$}}1
        {Φ}{{$\Phi$}}1
        {↾}{{$\upharpoonright$}}1
        {∀}{{$\forall$}}1
        {→}{{$\to$}}1
        {=>}{{$\Rightarrow$}}1
        {≫}{{$\gg$}}1
        {≃}{{$\simeq$}}1
}
\lstset{
    language=lean,
    breaklines=true,
    frame=single,
    showstringspaces=false
}

\title{Pointfree CoEndoYoneda Lemma \\ 
in \\
Functional Categories \\ 
Theory and Formalization in Lean 4}
\author{Luc Duponcheel \and Antigravity AI}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We present a variation of the Yoneda Lemma for functional categories equipped with a monadic
structure, which we formally identify as \emph{Functional Categories}. The central theoretical 
contribution, the Pointfree CoEndoYoneda Lemma, establishes a natural bijection between natural
transformations mapping out of a co-Yoneda endofunctor and corresponding global elements within the 
monadic functional category. In addition to outlining the mathematical foundations, we detail the
complete formalization of this result in the Lean 4 theorem prover. 

We discuss the technical challenges of working with functor compositions under strict definitional
equality and demonstrate strategies for navigating dependent type theory to construct fully formal,
pointfree proofs of higher-level category theory results.
\end{abstract}

\section{Introduction}

The \emph{CoYoneda Lemma} serves as a cornerstone of modern category theory, asserting that for any
category  $\mathcal{C}$, object $X \in \mathcal{C}$, and functor $F : \mathcal{C} \to \mathbf{Set}$,
the set of natural transformations $\text{Nat}(\mathcal{C}(X, -), F)$ is isomorphic to the
evaluation $F(X)$. This result fundamentally states that an object $X$ is completely determined by
the morphisms out of it.

In computational settings, we frequently encounter categories equipped with additional 
structure—specifically, categories that natively interpret types (or sets) and possess a monad-like
mechanism for handling effects. We define such a category as a \emph{Functional Category}. In this
paper, we explore how the Yoneda principle extends to this structural setting. 

The primary contribution is the \emph{CoEndoYoneda Lemma}, which characterizes natural
transformations that map out of an endofunctor defined via the co-Yoneda embedding. We state two
main lemmas, showing that such natural transformations are entirely determined by their evaluation
at the identity morphism. We have a type-level equivalence between natural transformations from the
endo-coyoneda embedding and global elements. The equivalence is, replacing elements by global
elements, entirely pointfree.

We also  provide a comprehensive discussion of how these concepts and theorems are
formalized in the Lean 4 proof assistant using the \texttt{Mathlib} library.

\section{Functional Categories}

Let $\mathcal{C}$ be a locally small category. We denote the category of types (or sets) as
$\mathbf{Type}$. For any object $X \in \mathcal{C}$, the covariant Hom-functor, which we will call
the co-Yoneda functor, is defined as:
\[
CY\!F_X : \mathcal{C} \to \mathbf{Type}, \quad Y \mapsto \mathcal{C}(X, Y)
\]

\begin{definition}
A \textbf{Functional Category} is a category $\mathcal{C}$ equipped with the following data:
\begin{enumerate}
    \item A designated functor $FF : \mathbf{Type} \to \mathcal{C}$.
    \item A constructed endofunctor $GF : \mathcal{C} \to \mathcal{C}$ 
          defined by \\
          $GF(X) = FF(CY\!F_{FF(1)}(X))$.
    \item A monadic structure over $GF$, provided by a \\
          unit natural transformation  $\gamma\eta : \text{Id}_{\mathcal{C}} \to GF$ and a \\
          multiplication natural transformation $\gamma\mu : GF \circ GF \to GF$.
\end{enumerate}

These structural components must satisfy the standard monad left unit law: 

\[
\gamma\eta_{GF(X)} \circ \gamma\mu_X = \text{id}_{GF(X)}
\]
\end{definition}

For now no other extra laws besides, of course, the naturality laws, are needed in the formal Lean 4
proof. This is an interesting observation on its own.

This structure essentially allows $\mathcal{C}$ to reflect the category of types into itself, while
treating $GF$ as a monadic context. To generalize Yoneda, we define the \emph{CoYoneda EndoFunctor}
$CY\!EF_X : \mathcal{C} \to \mathcal{C}$ for a fixed object $X$ as the composition:

\[
CY\!EF_X = CY\!F_X \circ FF
\]

We also explicitly define the concept of \emph{global elements} for an object $X \in \mathcal{C}$ as the morphisms from the canonical singleton $FF(1)$ to $X$:

\[
G_{\mathit{FF}}(X) = \mathcal{C}(FF(1), X)
\]

\subsection{Formalization of Functional Categories in Lean 4}

In Lean 4, this structure is elegantly captured using a type class extending the standard
\texttt{LargeCategory} from \texttt{Mathlib}. 

\newpage

\begin{lstlisting}[basicstyle=\tiny]
abbrev G {C : Type (u + 1)} [LargeCategory.{u} C] (T2C : Type u ⥤ C) (X : C) : Type u :=
  T2C.obj PUnit ⟶ X

class FunctionalCategory (C : Type (u + 1)) extends LargeCategory.{u} C where
  Φ : Type u ⥤ C

  γμ : (GEF_def Φ ⋙ GEF_def Φ) ⟶ GEF_def Φ
  γη : 𝟭 _ ⟶ GEF_def Φ

  γ_left_unit : 
    ∀ (X : C), γη.app ((GEF_def Φ).obj X) ≫ γμ.app X = 𝟙 ((GEF_def Φ).obj X)

  φ {X Y : Type u} (f : X → Y) : Φ.obj X ⟶ Φ.obj Y := Φ.map (TypeCat.ofHom f)

  φ_eq : 
    ∀ {X Y : Type u} (f : X → Y), φ f = Φ.map (TypeCat.ofHom f) := 
      by intros; rfl

  γη_φ :
    ∀ (Z : Type u), γη.app (Φ.obj Z) = φ (fun z => φ (fun _ => z)) :=
      by intros; rfl
\end{lstlisting}

This class natively encapsulates the natural transformations and the left unit law constraint. A
canonical instance of this class is $\mathbf{Type}$ itself, where $\Phi$ is the identity functor.

\begin{lstlisting}[basicstyle=\tiny]
instance typesFunctionalCategory : FunctionalCategory (Type u) where
  Φ := 𝟭 (Type u)
  γη := {
    app := fun X => (𝟭 (Type u)).map (↾(fun x => ↾(fun _ => x)))
    naturality := fun X Y f => by ext; rfl
  }
  γμ := {
    app := fun X => (𝟭 (Type u)).map (↾(fun (f : PUnit ⟶ PUnit ⟶ X) => ↾(fun pu => f pu pu)))
    naturality := fun X Y f => by ext; rfl
  }
  γ_left_unit := fun _ => by rfl
\end{lstlisting}

\section{The CoEndoYoneda Lemmas}

The core results consist of two main lemmas and a type-level equivalence showing how natural transformations out of $CY\!EF_Z$ uniquely
correspond to global elements.

\subsection{CoEndoYoneda Lemma 1}

\begin{lemma}
For any monadic functional category $\mathcal{C}$, endofunctor $F : \mathcal{C} \to \mathcal{C}$,
and object $Z \in \mathcal{C}$, there is an injection mapping natural transformations
$\tau : CY\!EF_Z \to F$ into transformations targeting $F \circ GEF$. 
\end{lemma}

We define \texttt{transformationToGlobalTransformation} which takes $\tau$ and post-composes it with the monad
unit $\gamma\eta$. Lemma 1 asserts that this transformation is identical to the one constructed by
taking the evaluation of $\tau$ at the identity morphism $1_Z$ and lifting it through $\Phi$.

\begin{lstlisting}[basicstyle=\tiny]
@[simps]
def transformationToGlobalTransformation {F : C ⥤ C} {Z : C} (τX : CYEF Z ⟶ F) :
    CYEF Z ⟶ (F ⋙ GEF) where
  app Y := τX.app Y ≫ γη.app (F.obj Y)

@[simps]
def globalValueToGlobalTransformation {F : C ⥤ C} {X : C} (gfx : Global (F.obj X)) :
    CYEF X ⟶ (F ⋙ GEF) where
  app Y := φ (gfx ≫ F.map .)

theorem coEndoYonedaLemma {F : C ⥤ C} {Z : C} (τX : CYEF Z ⟶ F) :
  transformationToGlobalTransformation τX =
    globalValueToGlobalTransformation (φ (fun _ => 𝟙 Z) ≫ τX.app Z)
\end{lstlisting}

\subsection{CoEndoYoneda Lemma 2}

Lemma 2 is the more substantial result, demonstrating a complete bijection when the target functor
is $F \circ GEF$. 

\begin{theorem}
Let $\tau : CY\!EF_Z \to F \circ GEF$ be a natural transformation. Then $\tau$ is uniquely determined
by its evaluation at $1_Z$. There exists a bijection between such transformations and global
elements $g \in G_{\Phi}(GEF(F(Z)))$.
\end{theorem}

In Lean, we construct the transformation out of the global element $g$ via \texttt{globalGlobalValueToGlobalTransformation} and then prove in \texttt{left\_inverse} that any
arbitrary $\tau$ is exactly this constructed transformation from its evaluated global value:

\begin{lstlisting}[basicstyle=\tiny]
def globalTransformationToGlobalGlobalValue {F : C ⥤ C} {X : C} (τX : CYEF X ⟶ F ⋙ GEF) :
  Global ((F ⋙ GEF).obj X) := φ (fun _ => 𝟙 X) ≫ τX.app X

@[simps]
def globalGlobalValueToGlobalTransformation
  {F : C ⥤ C} {X : C} (ggfx : Global (GEF.obj (F.obj X))) : CYEF X ⟶ (F ⋙ GEF) where
  app Y := φ (ggfx ≫ (F ⋙ GEF).map .) ≫ γμ.app (F.obj Y)

theorem left_inverse {F : C ⥤ C} {Z : C} (τX : CYEF Z ⟶ (F ⋙ GEF)) :
  τX =
    globalGlobalValueToGlobalTransformation (φ (fun _ => 𝟙 Z) ≫ τX.app Z)
\end{lstlisting}

\subsection{CoEndoYoneda Equivalence}

Building upon Lemma 2 and the dual inverse mapping, we establish a formal type-level equivalence between the set of natural 
transformations out of the co-Yoneda endofunctor and the corresponding global elements.
This result is the direct generalization of the classical CoYoneda equivalence to our monadic
functional category setting.

\begin{theorem}
There is a type-level equivalence between natural transformations $CY\!EF_X \to F \circ GEF$ 
and global elements $G_{\Phi}(GEF(F(X)))$.
\end{theorem}

In Lean 4, we define this as an \texttt{Equiv} (denoted by \texttt{$\simeq$}).

Utilizing \texttt{left\_inverse} provides the left inverse property.
For the right inverse, \texttt{right\_inverse} proves that evaluating the constructed transformation back at the identity yields the original global element, using the lemma \texttt{φ\_γμ} to collapse $\varphi(\text{const } ggfx) \gg \gamma\mu = ggfx$:

\begin{lstlisting}[basicstyle=\tiny]
theorem φ_γμ {X : C} (ggfx : Global ((GEF_def Φ).obj X)) :
    φ (fun _ => ggfx) ≫ γμ.app X = ggfx

theorem right_inverse {F : C ⥤ C} {X : C} :
  ∀ (ggfx : Global ((F ⋙ GEF).obj X)),
    ggfx =
      (fun τX => (φ fun _ => 𝟙 X) ≫ τX.app X) (globalGlobalValueToGlobalTransformation ggfx)

def coEndoYonedaEquiv {F : C ⥤ C} {X : C} : (CYEF X ⟶ (F ⋙ GEF)) ≃ Global ((F ⋙ GEF).obj X)
where
  toFun     := globalTransformationToGlobalGlobalValue
  invFun    := globalGlobalValueToGlobalTransformation
  left_inv  := fun τX => (left_inverse  τX).symm
  right_inv := fun τX => (right_inverse τX).symm
\end{lstlisting}

This equivalence provides a completely pointfree and logically verified dictionary for evaluating and constructing natural transformations in functional categories.

\section{Proof Techniques and Challenges in Lean 4}

Formalizing abstract category theory in dependent type theory (DTT) introduces unique difficulties,
particularly concerning definitional equality. 

\subsection{Strict Functor Composition}
In standard mathematics, functor composition is strictly associative. In Lean 4, however, these are
represented as distinct abstract syntax trees. They are \emph{isomorphic} via 
\texttt{Category.assoc}, but they are not \emph{definitionally equal}. 

When writing the proof of \texttt{left\_inverse}, this limitation prevents standard
term-rewriting tactics like \texttt{simp} from effortlessly reducing deeply nested compositions.

\subsection{Constructing Equality Chains}
To circumvent the strict type-checker without resorting to manual casts (such as \texttt{eqToHom}),
the proof for \texttt{left\_inverse} abandons the monolithic \texttt{simp} approach in favor of
explicit equality chains acting on the natural transformation components.

By utilizing \texttt{Eq.trans} and \texttt{congr\_arg}, we explicitly guide the type checker through
the unit laws and naturality squares:

\begin{lstlisting}[basicstyle=\tiny]
  have h1 : 
    τX.app Y = τX.app Y ≫ 𝟙 ((GEF_def Φ).obj (F.obj Y)) := 
      (Category.comp_id (τX.app Y)).symm
  have h2 : 
    𝟙 ((GEF_def Φ).obj (F.obj Y)) = 
      γη.app ((GEF_def Φ).obj (F.obj Y)) ≫ γμ.app (F.obj Y) :=
        (γ_left_unit (F.obj Y)).symm
  have h_left_unit : 
    τX.app Y = 
      τX.app Y ≫ (γη.app ((GEF_def Φ).obj (F.obj Y)) ≫ γμ.app (F.obj Y)) :=
        Eq.trans h1 (congr_arg (τX.app Y ≫ .) h2)
\end{lstlisting}

This localized approach bypasses the rigid typing of the $\longrightarrow$ category morphisms by
acting directly on the equality type \texttt{=} in the universe of types. This technique is highly
robust and avoids the infamous "motive is not type correct" errors common in Lean category theory
formalization.

\section{Conclusion}

The Pointfree CoEndoYoneda Lemma is a variation of the classical Pointful CoYoneda Lemma in a
monadic functional category setting. By successfully formalizing the lemma in Lean 4, we have not
only verified its absolute logical correctness but also established a concrete methodology for
handling strict composition equality in dependent type theory. The approach of combining explicit
structural definitions with targeted, step-by-step equality chaining provides a reliable method for
formalizing advanced category theory in modern proof assistants.

\end{document}
